Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-06-25 inst/doc/nonparametric-bayes.Rmd
writing

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.

Here we illustrate how a stochastic dynamic programming algorithm can 
be driven by the predictions from a Gaussian process model, sidestepping
the need for an accurate model-based description of the sustem dynamics.

<!-- Models for decision-making -->
Decision-theoretic (or optimal control) tools require a model that can assign probabilities of
future states (e.g. stock size of a fishery) given the current state and a
proposed action (e.g. fishing harvest or effort).  The decision maker then
seeks to determining the course of actions (also referred to as the policy)
that maximizes the expected value of some objective function (such as
net present value derived from the resource over time.  (The approach
can be adapted to permit alternatives to maximizing expectation, such
as minimizing the maximum cost or damage that might be incurred. See
@Polasky2011).
Management frequently faces a sequential decision-making problem -- after 
selecting an action, the decision-maker may receive new information about 
the current state and must again choose an appropriate action -- i.e. setting
the harvest limits each year based on stock assessments the year prior.

<!-- Parametric models -->

Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation.  Such simple 
parametric models, frequently justified by mechanistic underpinnings are used
to provide probabilities of future states.  Fisheries literature, which enjoys
some of the longest history in the use of the optimal control framework, provides
a clear example. Models such as the Ricker or Beverton Holt curves are ubiquitous [too strong? references?]
Parametric models can reflect variability introduced by a stochastic environment
or demographic process my incorporating random shocks from a given distribution.

<!-- Parametric uncertainty -->
As the parameter values for these models must be estimated from limited data,  
there will always be some uncertainty associated with these values.  This uncertainty
further compounds the intrinsic variability introduced by demographic or environmental 
noise.  The degree of uncertainty in the parameter values can be inferred from the data
and reflected in the estimates of the transition probabilities.

<!-- Structural Uncertainty --> 
Unfortunately, estimates of parameter uncertainty are only as good as the parametric 
models themselves.  Too often, we do not understand the system dynamics well enough
to know if a model provides a good approximation over the relevant range of states
and timescales (criteria that we loosely refer to as defining the "right" or "true" model.)
So called structural or model uncertainty is a more difficult problem than parametric 
uncertainty.  Typical solutions involve either model choice, or model averaging. Either
approach remains limited by the degree to which any of the proposed models are sufficiently
close to the right model.

<!-- Comparing models -->
The nature of decision-making problems provides a convenient way to compare 
models.  Rather than compare models in terms of best fit to data or fret over
the appropriate penalty for model complexity, model performance is defined 
in the concrete terms of the decision-maker's objective function, which we
will take as given. (Much argument can be made over the 'correct' objective
function, e.g. how to account for the social value of fish left in the sea
vs. the commercial value of fish harvested; see @Halpern2013 for further 
discussion of this issue.  Alternatively, we can always compare model performance
across multiple potential objective functions.)  The decision-maker
does not necessarily need a model that provides the best mechanistic understanding
or the best long-term outcome, but rather the one that best estimates the 
probabilities of being in different states as a result of the possible actions.

<!-- The weakness of parametric models -->

@@

We demonstrate how
a Gaussian Process model of stock recruitment can lead to nearly optimal
management through stochastic dynamic programming, comperable to knowing
the correct structural equation for the underlying simulation.  Meanwhile,
parametric models that do not match the underlying dynamics can perform 
very poorly, even though they fit the data as well as the true model.

Ecological research and management strategy should pay closer attention
to the opportunities and challenges nonparametric modeling can offer.

<!--
### Quantitative vs Qualitative Decisions

@@

### Terminology: "Non-parametric"

Approach and Methods
=====================================================================

To illustrate the value of the non-parametric Bayesian approach to management,
we focus on example of a system containing such a tipping point whose dynamics
can still be described by a simple, one-dimensional parametric model.  
We will focus on a simple parametric model for a single species [derived
from fist principles by @Allen2005a] as our underlying "reality".
X_{t+1} &= Z_t f(S_t) \\
S_t &= X_t - h_t \\
f(S_t) &= S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - C\right)}
\end{align}

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape parameter
$\sigma_g$.  We start with an example in which the parameters are $r =2$,
$K=8$, $C=5$ and $\sigma_g = 0.1$.
<!-- ` r p[1]`, $K =$ ` r p[2]`, $C =$ ` r p[3]`, and  $\sigma_g =$ ` r sigma_g`. -->

Both parametric and nonparametric approaches will require training
data on which to base their model of the process.  We generate the
training data under the model described in Eq 1 for $T_{\textrm{obs}}=40$ time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$.  For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the harvest.

<!-- things that we could hit upon in discussion:
Using data simulated from a specified model rather than empirical data
permits the comparison against the true underlying dynamics, setting 
a bar for the optimal performance possible.

(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).
-->

<span style="color:#007FFF;">***Should we include any emprical examples?***</span>

### The optimal control problem in fisheries management

@@

@@

We generate the data with a four-parameter model that contains
a tipping point (an Allee effect, see [@Allen, @Courchamp]) below which the stock
decreases to zero,

$$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K}\right)\left(\frac{S_t- \theta}{K}\right)} \label{Allen} $$

@@

$$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K} \right) } \label{Ricker} $$

@@

$$ X_{t+1} = Z_t \frac{r S_t^{\theta}}{1 - \frac{S_t^\theta}{K}} \label{Myers} $$

@@

@@

The essence of the GP approach can be captured in the
following thought experiment: An exhaustive parametric approach to the
challenge of structural uncertainty might proceed by writing down all
possible functional forms for the underlying dynamical system with all
possible parameter values for each form, and then consider searching
over this huge space to select the most likely model and parameters;
or using a Bayesian approach, assign priors to each of these possible
models and infer the posterior distribution of possible models. The
GP approach can be thought of as a computationally
efficient approximation to this approach. GPs represent
a large class of models that can be though of as capturing or reasonably
approximating the set of models in this collection.  By modeling at the
level of the process, rather than the level of parametric equation,
we can more concisely capture the possible behavior of these curves.
In place of a parametric model of the dynamical system, the GP 
approach postulates a prior distribution of (n-dimensional)
curves that can be though of as approximations to a range of possible
(parametric) models that might describe the data. The GP allows us
to consider a set of possible curves simultaneously.

### Background on Gaussian Process inference

@@

This completes the algorithm adapting the GP to the sequential decision-making 
problem through SDP, which we believe has not yet been demonstrated in the
literature.  We provide an R package implementation of this, along with
the Gaussian process inference, in the supplemental materials.

@@

The posterior distribution for the hyper-parameters of the Gaussian 
process model are estimated by Metropolis-Hastings algorithm, again with
details and code provided in the Appendix.  @Rasmussen2006 provides
an excellent general introduction to Gaussian Processes and @Munch2005 
first discusses their application in this context.

Results
====================================================

@@

Discussion 
==================================================================

@@

@@

## Future directions

@@

<span style="color:#007FFF;"> Data complexity. Perhaps too far out of scope... </span>
The nonparametric Bayesian approach is also better suited to complex and
disparate data.  Incorporating various sources of information into
mechanistic models can be an immensely difficult due to the increased
complexity involved.

@@

Acknowledgments
================

>>>>

#### Opening

[@Fischer2009, @Polasky2011]

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.  
Here we illustrate how a stochastic dynamic programming algorithm can 
be driven by the predictions from a Gaussian process model, sidestepping
the need for an accurate model-based description of the sustem dynamics.

#### Models for decision-making under uncertainty

Decision-theoretic (or optimal control) tools require a model that can assign probabilities of
future states (e.g. stock size of a fishery) given the current state and a
proposed action (e.g. fishing harvest or effort).  The decision maker then
seeks to determining the course of actions (also referred to as the policy)
that maximizes the expected value of some objective function (such as
net present value derived from the resource over time.  (The approach
can be adapted to permit alternatives to maximizing expectation, such
as minimizing the maximum cost or damage that might be incurred. See
@Polasky2011).
Management frequently faces a sequential decision-making problem -- after 
selecting an action, the decision-maker may receive new information about 
the current state and must again choose an appropriate action -- i.e. setting
the harvest limits each year based on stock assessments the year prior.

[@Reed1979, @Mangel1985]

#### Parametric models

Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation.  Such simple 
parametric models, frequently justified by mechanistic underpinnings are used
to provide probabilities of future states.  Fisheries literature, which enjoys
some of the longest history in the use of the optimal control framework, provides
a clear example [ ] . Models such as the Ricker or Beverton Holt curves are ubiquitous [too strong? references?]
Parametric models can reflect variability introduced by a stochastic environment
or demographic process my incorporating random shocks from a given distribution.

#### Parametric uncertainty

As the parameter values for these models must be estimated from limited data,  
there will always be some uncertainty associated with these values.  This uncertainty
further compounds the intrinsic variability introduced by demographic or environmental 
noise.  The degree of uncertainty in the parameter values can be inferred from the data
and reflected in the estimates of the transition probabilities.

[@Mangel1985; @Schapaugh2013]

#### Structural Uncertainty

Unfortunately, estimates of parameter uncertainty are only as good as the parametric 
models themselves.  Too often, we do not understand the system dynamics well enough
to know if a model provides a good approximation over the relevant range of states
and timescales (criteria that we loosely refer to as defining the "right" or "true" model.)
So called structural or model uncertainty is a more difficult problem than parametric 
uncertainty.  Typical solutions involve either model choice, or model averaging. Either
approach remains limited by the degree to which any of the proposed models are sufficiently
close to the right model.

[@Williams2001; @Athanassoglou2012].

#### Comparing models

The nature of decision-making problems provides a convenient way to compare 
models.  Rather than compare models in terms of best fit to data or fret over
the appropriate penalty for model complexity, model performance is defined 
in the concrete terms of the decision-maker's objective function, which we
will take as given. (Much argument can be made over the 'correct' objective
function, e.g. how to account for the social value of fish left in the sea
vs. the commercial value of fish harvested; see @Halpern2013 for further 
discussion of this issue.  Alternatively, we can always compare model performance
across multiple potential objective functions.)  The decision-maker
does not necessarily need a model that provides the best mechanistic understanding
or the best long-term outcome, but rather the one that best estimates the 
probabilities of being in different states as a result of the possible actions.

<!-- Transition and map: The weakness of parametric models -->
<!-- This section is a bit weak and vague.  Revisit to make a real map in light 
of outline structure below. -->

@@

We demonstrate how
a Gaussian Process model of stock recruitment can lead to nearly optimal
management through stochastic dynamic programming, comperable to knowing
the correct structural equation for the underlying simulation.  Meanwhile,
parametric models that do not match the underlying dynamics can perform 
very poorly, even though they fit the data as well as the true model.  
Ecological research and management strategy should pay closer attention
to the opportunities and challenges nonparametric modeling can offer.

#### Background on Fisheries Context

#### Background on Tipping points

<!-- note: This section largely replaced by above outline

### Quantitative vs Qualitative Decisions

@@

Approach and Methods
====================

## The optimal control problem in fisheries management

@@

To illustrate the value of the non-parametric Bayesian approach to management,
we focus on example of a system containing such a tipping point whose dynamics
can still be described by a simple, one-dimensional parametric model.  
We will focus on a simple parametric model for a single species [derived
from fist principles by @Allen2005a] as our underlying "reality".

$$X_{t+1} = Z_t f(S_t)  $$
$$S_t = X_t - h_t $$
$$f(S_t) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - C\right)} $$

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape parameter
$\sigma_g$.  We start with an example in which the parameters are $r =2$,
$K=8$, $C=5$ and $\sigma_g = 0.1$.

<!-- ` r p[1]`, $K =$ ` r p[2]`, $C =$ ` r p[3]`, and  $\sigma_g =$ ` r sigma_g`. -->

#### Simulated training data

We generate initial observational data under the model described in Eq 1 
for $T_{\textrm{obs}}=40$ time steps, under a given arbitrary sequence of harvest
intensities, $h_t$. 
We consider the case in which most of the data comes from a limited region
of state space (e.g. near a stable equilibrium), leaving us without
observations of the population dynamics at very low levels which would
be useful in discrimating between recruitment curves [@] or demonstrating
the existence of a tipping point [@Scheffer2001].  
Using data simulated from a specified model rather than empirical data
permits the comparison against the true underlying dynamics, setting 
a bar for the optimal performance possible.

@@

We generate the data with a four-parameter model that contains
a tipping point, as discussed above (equation 1),
(an Allee effect, see [@Allen, @Courchamp]) below which the stock
decreases to zero,

$$ X_{t+1} = Z_t S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(\frac{S_t - \theta}{K}\right)} $$

@@

$$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K} \right) } $$

@@

$$ X_{t+1} = Z_t \frac{r S_t^{\theta}}{1 - \frac{S_t^\theta}{K}} $$

@@

#### Terminology: "Non-parametric"

@@

The essence of the GP approach can be captured in the
following thought experiment: An exhaustive parametric approach to the
challenge of structural uncertainty might proceed by writing down all
possible functional forms for the underlying dynamical system with all
possible parameter values for each form, and then consider searching
over this huge space to select the most likely model and parameters;
or using a Bayesian approach, assign priors to each of these possible
models and infer the posterior distribution of possible models. The
GP approach can be thought of as a computationally
efficient approximation to this approach. GPs represent
a large class of models that can be though of as capturing or reasonably
approximating the set of models in this collection.  By modeling at the
level of the process, rather than the level of parametric equation,
we can more concisely capture the possible behavior of these curves.
In place of a parametric model of the dynamical system, the GP 
approach postulates a prior distribution of (n-dimensional)
curves that can be though of as approximations to a range of possible
(parametric) models that might describe the data. The GP allows us
to consider probabilities on a large set of possible curves simultaneously.

The posterior distribution for the hyper-parameters of the Gaussian 
process model are estimated by Metropolis-Hastings algorithm, again with
details and code provided in the Appendix.  @Rasmussen2006 provides
an excellent general introduction to Gaussian Processes and @Munch2005 
first discusses their application in the context of population dynamics
models such as fisheries stock-recruitment relationships.

### SDP via GP

@@

This completes the algorithm adapting the GP to the sequential decision-making 
problem through SDP, which has not been previously demonstrated.  
We further provide an R package implementation as described in the supplemental materials.

@@

Results
=======

@@

Discussion 
==========

#### Big picture: Linking GP to SDP

@@

<!-- On mechanistic models -->

@@

Future directions
-----------------

### Higher dimensions

@@

@@

Acknowledgments
===============
